Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-14 prop_letter.tex
prop_letter.tex \section{Gleason} Updating...

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$, i.e., $q_{j} = j$ and
$\ket{\phi_{j}} = \ket{j - 1}$.
Two fragments of valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the
second corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. \yutsung{Notice that we also have \mu_{1}=\mu_{\rho}^{\mathrm{B}}
for a mixed state \rho=0.5\proj{0}+0.5\proj{1}. Maybe we should add ``Or other mixed states.''} 
The expectation values of the observable $\mathbf{O}$,
$\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the first case and 2 in
the second.

@@

The quantum probability measure~$\mu_{1}$ maps some projectors to
$0$ or $0.5$, which are replaced by $\left[0,0\right]$ or $\left[0.5,0.5\right]$
to get $\bar{\mu}_{3}$. This is replacing two numbers by the smallest
intervals containing them. In general, any set of the numbers can
be replaced by the interval containing them if we carefully pick a
set of overlapping intervals. For example, consider the following
collection of intervals: 
\begin{eqnarray}
\widehat{\mathscr{I}}_{n} & = & \set{\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\{\imposs,\necess\}
\end{eqnarray}
which are overlapping intervals with length~$1/n$ except the intervals
at length~$\left(2n\right)^{-1}$ at each end. Then, we can define
a QIVPM mapping to $\widehat{\mathscr{I}}_{n}$ based on any Born
rule probability~$\mu_{\rho}^{\mathrm{B}}$ on a three-dimensional
Hilbert space by replacing the probability~$\mu_{\rho}^{\mathrm{B}}\left(P\right)$
with the interval containing it, $\widehat{\iota}_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$.
Since intervals in $\widehat{\mathscr{I}}_{n}$ are overlapping, if
$\mu_{\rho}^{\mathrm{B}}\left(P\right)$ are contained in both intervals~$[\ell_{1},r_{1}]$
and $[\ell_{2},r_{2}]$, $\widehat{\iota}_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
will be $[\ell_{i},r_{i}]$ if $\widehat{\iota}_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
is closer to the center of $[\ell_{i},r_{i}]$\yutsung{{Verify...}}.
When there is a tie, we set $\widehat{\iota}_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)=\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]$
for even $k$. For example, when $n=2$, 
\begin{eqnarray}
\widehat{\mathscr{I}}_{2} & = & \set{\left[\frac{k-1}{4},\frac{k+1}{4}\right]\cap\left[0,1\right]}{k=0,\ldots,4}\nonumber \\
 &  & \cup\{\imposs,\necess\}\nonumber \\
 & = & \left\{ \imposs,\left[0,\frac{1}{4}\right],\left[0,\frac{1}{2}\right],\left[\frac{1}{4},\frac{3}{4}\right],\right.\nonumber \\
 &  & \left.\left[\frac{1}{2},1\right],\left[\frac{1}{4},1\right],\necess\right\} \,.
\end{eqnarray}
It is straightforward to verify $\widehat{\iota}_{2}\circ\mu_{1}$
and $\widehat{\iota}_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$ are
QIVPMs whose values on the minimal subspace of events containing $\proj{0}$,
$\proj{1}$, and $\proj{2}$, $\events_{C}$, is computed by Table~\ref{tab:Compute-the-QIVPM}.
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\widehat{\iota}_{2}\circ\mu_{1}$
and $\widehat{\iota}_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$ based
on $\mu_{1}$ and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$ on $\events_{C}$.}
\begin{tabular}{ccccc}
\toprule 
$P$ & $\mu_{1}\left(P\right)$ & $\widehat{\iota}_{2}\left(\mu_{1}\left(P\right)\right)$ & $\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)$ & $\widehat{\iota}_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline
\midrule
\midrule 
$\mathbb{0}$ & $0$ & $\left[0,0\right]$ & $0$ & $\left[0,0\right]$\tabularnewline
\midrule 
$\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{2}$ & $0$ & $\left[0,0\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{2}$ & $1$ & $\left[1,1\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}$ & $1$ & $\left[1,1\right]$ & $1$ & $\left[1,1\right]$\tabularnewline
\bottomrule
\end{tabular}
\end{table}

Given any state~$\rho$ satisfying $0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$
for any $P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $,
called totally mixed state, we know not only $\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is a QIVPM, but also the collection of states consistent it get closer
and closer to $\rho$ as the intervals get narrower and narrower,
eventually collapsing to $\rho$ when the intervals become infinitely
precise real numbers. To state this result, we need to define a notion
of distance based on the following norm of a matrix~$A$ as $\left\Vert A\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$~\citep{544199}.
If a state~$\rho'$ is consistent with $\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
its distance with $\rho$, $\left\Vert \rho-\rho'\right\Vert $, must
be bounded by $1/\left(12n\right)$. Reversely, if the distance between
$\rho$ and $\rho'$ is bounded by $1/n$, then $\rho'$ is consistent
with $\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$. In other
words, for the states consistent with this QIVPM, $\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho'$ as $n$ goes to infinity.

>>>>

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$, i.e., $q_{j} = j$ and
$\ket{\phi_{j}} = \ket{j - 1}$.
Two fragments of valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the
second corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. \yutsung{Notice that we also have
$\mu_{1}=\mu_{\rho}^{\mathrm{B}}$ for a mixed state~$\rho=0.5\proj{0}+0.5\proj{1}$.
Maybe we should add ``Or other mixed states.''}
The expectation values of the observable $\mathbf{O}$,
$\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the first case and 2 in
the second.

@@

The quantum probability measure~$\mu_{1}$ maps some projectors to
$0$ or $0.5$, which are replaced by $\left[0,0\right]$ or $\left[0.5,0.5\right]$
to get $\bar{\mu}_{3}$. This is replacing two numbers by the smallest
intervals containing them. In general, any set of the numbers can
be replaced by an interval containing them if we carefully pick a
set of intervals and the rule to select the correct interval when
intervals are overlapping. For example, consider the following collection
of intervals: 
\begin{eqnarray}
\mathscr{I}_{n} & = & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\{\imposs,\necess\}
\end{eqnarray}
which are overlapping intervals with length~$1/n$ except the intervals
at length~$\left(2n\right)^{-1}$ at each end. Since these intervals
are overlapping, we define the interior part of these intervals

Then, we can define a QIVPM mapping to $\mathscr{I}_{n}$ based on
any Born rule probability~$\mu_{\rho}^{\mathrm{B}}$ on a three-dimensional
Hilbert space by replacing the probability~$\mu_{\rho}^{\mathrm{B}}\left(P\right)$
with the interval containing it, $\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$.
Since intervals in $\mathscr{I}_{n}$ are overlapping, if $\mu_{\rho}^{\mathrm{B}}\left(P\right)$
are contained in both intervals~$[\ell_{1},r_{1}]$ and $[\ell_{2},r_{2}]$,
$\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$ will
be $[\ell_{i},r_{i}]$ if $\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
is closer to the center of $[\ell_{i},r_{i}]$\yutsung{Verify...}.
When there is a tie, we set $\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)=\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]$
for even $k$. For example, when $n=2$, 
\begin{eqnarray}
\mathscr{I}_{2} & = & \set{\left[\tfrac{k-1}{4},\tfrac{k+1}{4}\right]\cap\left[0,1\right]}{k=0,\ldots,4}\nonumber \\
 &  & \cup\{\imposs,\necess\}\nonumber \\
 & = & \left\{ \imposs,\left[0,\tfrac{1}{4}\right],\left[0,{\textstyle \tfrac{1}{2}}\right],\left[\tfrac{1}{4},\tfrac{3}{4}\right],\right.\nonumber \\
 &  & \left.\left[\tfrac{1}{2},1\right],\left[\tfrac{1}{4},1\right],\necess\right\} \,.
\end{eqnarray}
It is straightforward to verify $\iota_{2}\circ\mu_{1}$ and $\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$
are QIVPMs whose values on the minimal subspace of events containing
$\proj{0}$, $\proj{1}$, and $\proj{2}$, $\events_{C}$, is computed
by Table~\ref{tab:Compute-the-QIVPM}. 
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\iota_{2}\circ\mu_{1}$ and
$\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$ based on $\mu_{1}$
and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$ on $\events_{C}$.}
\begin{tabular}{ccccc}
\toprule 
$P$ & $\mu_{1}\left(P\right)$ & $\iota_{2}\left(\mu_{1}\left(P\right)\right)$ & $\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline
\midrule
\midrule 
$\mathbb{0}$ & $0$ & $\left[0,0\right]$ & $0$ & $\left[0,0\right]$\tabularnewline
\midrule 
$\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{2}$ & $0$ & $\left[0,0\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{2}$ & $1$ & $\left[1,1\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}$ & $1$ & $\left[1,1\right]$ & $1$ & $\left[1,1\right]$\tabularnewline
\bottomrule
\end{tabular}
\end{table}

Given any state~$\rho$ satisfying $0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$
for any $P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $,
called totally mixed state, we know not only $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is a QIVPM, but also the collection of states consistent it get closer
and closer to $\rho$ as the intervals get narrower and narrower,
eventually collapsing to $\rho$ when the intervals become infinitely
precise real numbers. To state this result, we need to define a notion
of distance based on the following norm of a matrix~$A$ as $\left\Vert A\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$~\citep{544199}.
If a state~$\rho'$ is consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
its distance with $\rho$, $\left\Vert \rho-\rho'\right\Vert $, must
be bounded by $1/\left(12n\right)$. Reversely, if the distance between
$\rho$ and $\rho'$ is bounded by $1/n$, then $\rho'$ is consistent
with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$. In other words, for
the states consistent with this QIVPM, $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho'$ as $n$ goes to infinity.
